XRONOS: an open data infrastructure for archaeological chronology

### 2024-12-02 paper.qmd
Add fig-c14-global

---
title: "XRONOS: an open data infrastructure for archaeological chronology"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    email: joseph.a.roe@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    country: Switzerland
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    country: Germany
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It aims provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world. By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records. It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating. Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it significantly easier to filter data of interest and reducing the need for manual data cleaning by end users. In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
keywords: 
  - open data
  - chronology
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
execute:
  echo: false
number-sections: true
bibliography: references.bib
format:
  hikmah-pdf:
    papersize: a4
    mainfont: "Linux Libertine O" # had to adjust, otherwise was not rendering
    sansfont: "URW Gothic"
    biblatex-chicago: true
    biblio-style: authordate
    biblatexoptions:
      - backend=biber
      - autolang=hyphen
      - isbn=false
      - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
---

```{r setup}
#| include: false
library("countrycode")
library("cowplot")
library("dplyr", warn.conflicts = FALSE)
library("dm")
library("giscoR")
library("ggplot2")
library("glue")
library("gt", warn.conflicts = FALSE)
library("khroma")
library("magick")
library("patchwork", warn.conflicts = FALSE)
library("readr")
library("RPostgres")
library("sf")
library("tidyr")
library("webshot2")

@@

In addition, the temporal and geographic coverage of these resources is uneven [@AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative of each other.
<!-- TODO: example, X databases for Europe, Y for the rest of the world? -->
By our count, <!-- TODO: X --> of the databases are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @OpenKnowledgeFoundation], which both limits the access to and reuse potential of these datasets.
<!-- TODO: X --> are not currently available in readily machine-readable formats (e.g. plain text or database files rather than PDFs or hypertext).

@@

<!-- TODO: Include CalPal?? -->

@@

The first available synthetic radiocarbon database was c14bazAAR [@SchmidEtAl2019], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (@FIG).

@@

As of writing, c14bazAAR includes <!-- TODO: N --> radiocarbon dates with unique laboratory identifiers, IntChron <!-- TODO: N --> (excluding those from non-archaeological contexts), and p3k14c <!-- TODO: N -->.
The geographic distribution of dates from each is similar (@FIG), reflecting the large degree in overlap between the sources of c14bazAAR and p3k14c in particular.
IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but is much smaller.

@@

# Dear god why
xronos_dm_svg_file <- tempfile("xronos_dm_", fileext = ".svg")
write(xronos_dm_svg, xronos_dm_svg_file)
ggdraw() + draw_image(xronos_dm_svg_file)
```

>>>>

---
title: "XRONOS: an open data infrastructure for archaeological chronology"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    email: joseph.a.roe@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    country: Switzerland
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    country: Germany
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It aims provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world. By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records. It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating. Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it significantly easier to filter data of interest and reducing the need for manual data cleaning by end users. In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
keywords: 
  - open data
  - chronology
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
number-sections: true
bibliography: references.bib
format:
  hikmah-pdf:
    papersize: a4
    mainfont: "Linux Libertine O" # had to adjust, otherwise was not rendering
    sansfont: "URW Gothic"
    biblatex-chicago: true
    biblio-style: authordate
    biblatexoptions:
      - backend=biber
      - autolang=hyphen
      - isbn=false
      - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
execute:
  echo: false
  cache: true
---

```{r setup}
#| include: false
library("countrycode")
library("cowplot")
library("dplyr", warn.conflicts = FALSE)
library("dm")
library("giscoR")
library("ggplot2")
library("glue")
library("gt", warn.conflicts = FALSE)
library("here")
library("khroma")
library("magick")
library("patchwork", warn.conflicts = FALSE)
library("purrr")
library("readr")
library("RPostgres")
library("sf")
library("spatstat")
library("stars")
library("tidyr")
library("webshot2")

@@

In addition, the temporal and geographic coverage of these resources is uneven [@ChaputGajewski2016; @AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative of each other.
<!-- TODO: example, X databases for Europe, Y for the rest of the world? -->
By our count, <!-- TODO: X --> of the databases are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @OpenKnowledgeFoundation], which both limits the access to and reuse potential of these datasets.
<!-- TODO: X --> are not currently available in readily machine-readable formats (e.g. plain text or database files rather than PDFs or hypertext).

@@

```{r data-fig-c14-global, include=FALSE}
c14_global <- read_tsv(here("data", "c14_global.tsv"))
c14_global_sum <- read_tsv(here("data", "c14_global_sum.tsv"))

ppp_window <- st_transform(st_union(gisco_get_countries()), "+proj=moll")

c14_global_density <- c14_global |>
  drop_na(longitude, latitude) |>
  nest(.by = source) |>
  mutate(
    sf = map(data, st_as_sf, coords = c("longitude", "latitude"), crs = 4326),
    sf = map(sf, st_transform, crs = "+proj=moll"),
    sf = map(sf, st_filter, ppp_window), # no dates in the sea, please
    n_sf = map_int(sf, nrow),
    ppp = map(sf, as.ppp, W = as.owin(ppp_window)),
    density = map(ppp, density, sigma = 50000, adjust = 10, eps = 50000),
    density = map(density, st_as_stars)
  )
```

```{r fig-c14-global, message=FALSE, warning=FALSE}
#| fig-cap: Geographic and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
#| fig-height: 6
plot_c14_density <- function(density, title, coastline = coast) {
  ggplot() +
    geom_stars(
      aes(fill = ecdf(v)(v)), # transform density to quantiles like plot.stars()
      data = density
    ) +
    geom_sf(data = coastline, fill = NA) +
    scale_fill_stepsn(
      name = "Density",
      n.breaks = 11,
      colours = colour("turku", reverse = TRUE)(11),
      labels = c("Low", rep("", 7), "High"),
      na.value = NA,
      guide = guide_none()
    ) +
    labs(title = title) +
    coord_sf(crs = "+proj=moll") +
    theme_minimal_grid() +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.margin = unit(c(11, 5.5, 0, 5.5), "pt")
    )
}

fig_density <- map2(c14_global_density$density, c14_global_density$source, 
                    plot_c14_density)
names(fig_density) <- c14_global_density$source

# From https://stackoverflow.com/questions/11053899/how-to-get-a-reversed-log10-scale-in-ggplot2
transform_reverse_log <- function(base = exp(1)) {
  trans <- function(x) -log(x, base)
  inv <- function(x) base^(-x)
  scales::trans_new(paste0("reverselog-", format(base)), trans, inv, 
                    scales::log_breaks(base = base), 
                    domain = c(1e-100, Inf))
}

plot_c14_sum <- function(data, n) {
  ggplot(data, aes(age, pdens)) +
    geom_area(fill = "#3C3B39") + # 'XRONOS grey'
    scale_x_continuous(name = NULL, breaks = c(50000, 5000, 500),
                       labels = c("50000", "5000", "500 BP"),
                       transform = transform_reverse_log()) +
    scale_y_continuous(name = NULL, labels = NULL) +
    labs(caption = paste0("N=", n)) +
    theme_minimal_vgrid() + 
    theme(
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = unit(c(0, 5.5, 11, 5.5), "pt")
    )
}

fig_sum <- c14_global_sum |>
  nest(.by = source) |>
  pull(data) |>
  map2(c14_global |> nest(.by = source) |> pull(data) |> map(nrow), plot_c14_sum)
names(fig_sum) <- unique(c14_global_sum$source)

# Lay out with patchwork
fig_density[["c14bazAAR"]] + fig_density[["IntChron"]] +
  fig_sum[["c14bazAAR"]] + fig_sum[["IntChron"]] +
  fig_density[["p3k14c"]] + fig_density[["XRONOS"]] +
  fig_sum[["p3k14c"]] + fig_sum[["XRONOS"]] +
  plot_layout(ncol = 2, heights = rep(c(5, 1), 2))

# Alt. plot by country
# c14_global |>
#   count(country, source) |>
#   left_join(gisco_get_countries(), by = c("country" = "CNTR_ID")) |>
#   mutate(n = replace_na(n, 0)) |>
#   st_as_sf() |>
#   ggplot() +
#   facet_wrap(vars(source)) +
#   geom_sf(data = gisco_get_countries(), fill = "white", colour = NA) +
#   geom_sf(aes(fill = n), colour = NA) +
#   geom_sf(data = coast, fill = NA) +
#   scale_fill_turku(reverse = TRUE, labels = scales::comma) +
#   labs(fill = "Dates") +
#   coord_sf(crs = "+proj=moll") +
#   theme_minimal_grid()
```

@@

The first available synthetic radiocarbon database was c14bazAAR [@SchmidEtAl2019], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
Despite this limitation, it has global coverage and a large number of dates (@fig-c14-global), and was therefore our starting point for data collection for XRONOS.

@@

As of writing, c14bazAAR has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")` radiocarbon dates with unique laboratory identifiers (excluding those sourced from p3k14c and XRONOS), IntChron has `r format(nrow(filter(c14_global, source == "IntChron")), big.mark = ",")` (excluding those from non-archaeological contexts), and p3k14c has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")`.
The geographic distribution of dates from each is similar (@fig-c14-global), reflecting the large degree in overlap between the sources of each compilation.
IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but less overall.

@@

# Dear god why is this necessary
xronos_dm_svg_file <- tempfile("xronos_dm_", fileext = ".svg")
write(xronos_dm_svg, xronos_dm_svg_file)
ggdraw() + draw_image(xronos_dm_svg_file)
```
